---
title: "Vectorized R code"
author: "Brett Melbourne"
date: "2022-10-14"
output:
    github_document
---

This program illustrates the difference between using repetition and selection structures versus vectorized operations.

Vectorized code is often much faster in R and many similar environments (e.g. Matlab). In some cases vectorized code is more natural to read. In other cases vectorized code can be more obtuse. There are pros and cons.


## Example 1

Simple addition of two vectors.

```{r}
A <- c(1,  4, -2, 3, 8)
B <- c(1, -8, -3, 6, 5)
```
```{r}
A
```
```{r}
B
```

To add A and B, we could do this:

```{r}
C <- rep(NA, 5)
for ( i in 1:5 ) {
    C[i] <- A[i] + B[i]
}
C
```

But in R, that would be nuts! We can vectorize this operation. Just do this:

```{r}
C <- A + B
C
```

It's doing exactly the same thing as the previous R `for` loop but the `for` loop is now behind the scenes in much faster, compiled, C code. The `+` in R is actually a function.

```{r}
"+"(A, B)
```

and if you delve deep enough in the source code of the function, you will ultimately find C code like this (slightly edited for clarity):

```{c, eval=FALSE}
double plus( double s1, double s2, int n ) {

    double ans[n];

    for ( int i = 0; i < n; i = i + 1 ) {
        ans[i] = s1[i] + s2[i];
    }

    return ans;
}
```

There isn't actually a function called "plus". This is the case `PLUSOP` in the function `real_binary` in the source code file `arithmetic.c` in `\src\main`.


## Example 2

This example contrasts two methods for calculating the vector y from the vector x for the linear model y[i] = b_0 + b_1 x[i], where i indexes the elements of the vectors. The function `system.time()` is used to time the calculations for comparison.

```{r}
vector_length <- 1e8	    #A large vector
x <- runif(vector_length)	#Create some values to use for x
b_0 <- 1.2
b_1 <- 2.5
```

For loop, writing to elements of the new vector, takes about 7 seconds on a 1.7 GHz Intel i5 processor (2019).

```{r}
system.time(
{
	y <- rep(NA, length(x))
	for ( i in 1:length(x) ) {
		y[i] <- b_0 + b_1 * x[i]
	}
}
)
```

R's native vectorized operation takes about 0.3 seconds, about 40X faster. The code is much simpler too!

```{r}
system.time(
	y <- b_0 + b_1 * x
)
```

As we saw above, behind the scenes R takes these instructions (the R code) and dispatches the work to the R source code that is written in C. This C code uses C `for` loops to do the addition and multiplication operations. These C `for` loops have the same algorithmic structure as the slower R code algorithm above but it runs super fast in C.


## Example 3

Logical and comparison operations can also be vectorized, often eliminating both repetition and selection structures at the same time. Here is an example like the student grades question in the selection structures assignment. The goal is to record "Passed" or "Failed" with a threshold grade of 60 for a class of students. I set the number of students here to a very high number to illustrate the difference in compute time.

Read in the data (generated here for illustration)

```{r}
numStudents <- 1e7		
grades <- runif(numStudents, 0, 100)
```

For loop, writing to a result vector. About 2.2 seconds.

```{r}
system.time(
{
	result <- rep(NA, numStudents)
	for ( i in 1:numStudents ) {
		if ( grades[i] < 60 ) {
			result[i] <- "Failed"
		} else {
			result[i] <- "Passed"
		}
	}
}
)
```

You can use fundamental vectorized operations to achieve the same thing. This is about 7X faster.

```{r}
system.time(
{
    result <- rep(NA, numStudents)
    result[ ( grades < 60 ) ] <- "Failed"
	result[ !( grades < 60 ) ] <- "Passed"
}
)
```

Faster still is this version (but it might accidentally default to passing some students if there are issues in the data)

```{r}
system.time(
{
    result <- rep("Passed", numStudents)
    result[ ( grades < 60 ) ] <- "Failed"
}
)
```

A neat vectorized solution is provided by the `ifelse()` function. This requires only one line of code but it takes even longer than the `for` loop. So, vectorizing is *not always* faster than `for` loops.

```{r}
system.time(
	result <- ifelse( grades < 60, "Failed", "Passed" )
)
```
