---
title: Quantiles and Q-Q plots
author: Brett Melbourne
date: 6 Oct 2021 (updated 17 Oct 2022)
output: github_document
---

A short tutorial on quantiles and Q-Q plots


### Understanding quantiles

Here is the cumulative distribution function (cdf) for the standard Normal distribution. This is a plot of the cumulative probability, Prob(X < x), versus x.

```{r}
x <- seq(-3, 3, length.out=100)
plot(x, pnorm(x), type="l", col="red", main="Cumulative distribution function")
axis(2, at=seq(0, 1, by=0.1), labels=FALSE, tcl=-0.25)
```

Quantiles are the inverse of this function. What is the value of x for which the cumulative probability is p? For example, what value of x are 80% of the values expected to be less than?

```{r}
p <- 0.80
qnorm(p) #This is that value of x; it equals 0.8416.
```

We say that 0.8416 is the quantile for p = 0.8. 

We could alternatively read the quantile off the graph of the cdf

```{r}
plot(x, pnorm(x), type="l", col="red", main="Cumulative distribution function")
axis(2, at=seq(0, 1, by=0.1), labels=FALSE, tcl=-0.25)
segments(-4, p, qnorm(p), p, col="blue")
segments(qnorm(p), -0.1, qnorm(p), p, col="blue")
text(qnorm(p), 0, labels=signif(qnorm(p), 4), pos=4)
```

Let's check that. Let's simulate a dataset from a standard normal (i.e. mean = 0, sigma = 1).

```{r}
z <- rnorm(100000)
```

Are 80% of the data less than the 0.8 quantile of 0.8416?

```{r}
sum(z < 0.8416) / 100000
```

Yes, that's about right.


### Procedure for constructing a QQ plot

The basic idea of a QQ plot is to plot the empirical quantiles (i.e. from the data) against the theoretical quantiles of the distribution in question (e.g. Normal, Poisson, etc).

Let's say we have some data, y. We'll imagine these are real data from now on. It's a small dataset, say 20 points.

```{r}
set.seed(23)    #to reproduce the exact example
y <- rnorm(20)
```

First of all, the data are their own quantiles, if we order them.

```{r}
q <- sort(y)
q #The observed quantiles
```

So, for example, the fourth value of the ordered data is the quantile with p = 4/20. That is 4/20 of the data are less than or equal to this value (but we'll take that as an estimate for strictly less than):

```{r}
q[4]
```

Now, because this is a finite sample and the sample is quite small and we won't expect the quantile for p = 1.0 to be in the sample, we include a small adjustment to better approximate the probabilities.

```{r}
n <- length(y)          #Number of data points
p <- ( 1:n - 0.5 ) / n  #We'll take these as the probabilities
cbind(p, q)             #The probabilities with their quantiles
```

The relationship between p and the ordered y (i.e. q) is the empirical cdf

```{r}
plot(q, p, type="l", col="red", main="Empirical cdf (red line)")
axis(2, at=seq(0, 1, by=0.1), labels=FALSE, tcl=-0.25)
for ( i in 1:n ) {
    segments(-4, p[i], q[i], p[i], col="blue")
    segments(q[i], -0.1, q[i], p[i], col="blue")    
}
```

Now we calculate the quantiles of the standard normal for the same probabilities (p) as the data

```{r}
qth <- qnorm(p)
qth #The theoretical quantiles
```

And we can plot those on the theoretical cdf

```{r}
plot(x, pnorm(x), type="l", col="red", ylab="p",
     main="Theoretical cdf (red curve)")
axis(2, at=seq(0, 1, by=0.1), labels=FALSE, tcl=-0.25)
for ( i in 1:n ) {
    segments(-4, p[i], qth[i], p[i], col="blue")
    segments(qth[i], -0.1, qth[i], p[i], col="blue")    
}
```

The Q-Q plot is the observed quantiles versus the theoretical quantiles. If the data are well approximated by the theoretical distribution (in this case Normal), then this should be a straightish line. Since the dataset is small it won't be perfectly straight.

```{r}
plot(qth, q)
```

The following code makes the same plot using R's built in functions and adds a line to indicate theoretically where the points should be.

```{r}
qqnorm(y)
qqline(y)
```

Explore how the Q-Q plot varies depending on the data. You can try running the following code several times and see how much it differs. Or try simulating a larger dataset. Or try simulating a dataset that is Normal with different mean and sigma (the line will still be straight), or generate data from a non-Normal distribution, such as lognormal (`rlnorm()`) or Poisson (`rpois()`) (the line will not be straight unless the distribution happens to be approximated well by the Normal).

```{r, eval=FALSE}
y <- rnorm(20, mean=0, sd=1)
qqnorm(y)
qqline(y)
```
