---
title: Ant data GLM - priors for Bayesian model
author: Brett Melbourne
date: 31 Oct 2022
output: github_document
---

Here we look at the priors in the GLM.

```{r results=FALSE, message=FALSE, warning=FALSE}
library(ggplot2)
library(rstanarm)
options(mc.cores=parallel::detectCores())
theme_set(theme_bw())
```

Read in data and train the model with `rstanarm`

```{r}
ant <- read.csv("data/ants.csv")
ant$habitat <- factor(ant$habitat)
bysfitHxL <- stan_glm(richness ~ habitat + latitude + habitat:latitude, 
                      family=poisson, data=ant)
```

To see what priors have been used

```{r}
prior_summary(bysfitHxL)
```

Before `rstanarm` trains the model it centers and scales the predictor variables. This improves the behavior of the MCMC algorithm resulting in faster and more reliable convergence. The default priors are specified on this scale. For this model (GLM, Poisson with log link) we see that the priors for the intercept and the three betas are all Normal(0,2.5): `location` is the mean of the distribution and `scale` is the standard deviation. There is a corresponding `Adjusted prior`, which is the equivalent prior on the original scale.

What does a Normal(0,2.5) prior look like? To visualize, we can compute it:

```{r}
# Calculate the curve for Normal(0,2.5)
theta <- seq(-10, 10, 0.1)
prior2.5 <- dnorm(theta, mean=0, sd=2.5)
# Outline polygon for 1 sd area under the curve
auc_1_sd <- c(prior2.5[abs(theta) <= 1], 0, 0)
theta_uc <- c(theta[abs(theta) <= 1], 1, -1)
auc_df <- data.frame(theta_uc, auc_1_sd)
# Plot distribution and 1 sd under curve
data.frame(theta, prior2.5) |>
    ggplot() +
    geom_polygon(data=auc_df, 
                 mapping=aes(x=theta_uc, y=auc_1_sd), fill="grey") +
    geom_line(mapping=aes(x=theta, y=prior2.5)) +
    ylab("density")
```

This is a weakly informative prior. We are expressing a tendency for the parameter to be near zero but most of the probability mass (area under the curve) is beyond 1 standard deviation unit, allowing plausible values out to about 6 or 7 standard deviation units. But magnitudes as large as 10 standard deviation units are very implausible.